---
title: "Reproducible Research - Peer Assessment 1"
author: "Amy White"
date: "5/14/2019"
output: html_document
---

```{r, message=FALSE}
library(lubridate)
library(ggplot2)
library(dplyr)
```

## Data preparation

Loading data.  Set cache=TRUE so only runs once.
```{r}
df <- read.csv("activity.csv",stringsAsFactors = F)
head(df)
summary(df)
```

Reformatting date variable
```{r}
df$date <- ymd(df$date)
class(df$date)
```

## Data analysis

Checking mean number of steps per day, ignoring missing values
```{r}
total_daily_steps <- df %>%
      group_by(date) %>%
      summarize(total_steps = sum(steps,na.rm=T))
hist(total_daily_steps$total_steps,main="Histogram of total daily steps")

mean_total_daily_steps <- 
      round(mean(total_daily_steps$total_steps),2)
median_total_daily_steps <- 
      round(median(total_daily_steps$total_steps),2)
```
The mean total daily steps is `r mean_total_daily_steps` and the median is `r median_total_daily_steps`.


Checking how average step counts change by 5 minute interval
```{r}
mean_interval_steps <- df %>%
      group_by(interval) %>%
      summarize(mean_steps = mean(steps,na.rm=T))
with(mean_interval_steps, plot(interval,mean_steps,
                              type="l",
                              main="Plot of Average Steps per Time Interval"))
max_interval <- mean_interval_steps[mean_interval_steps$mean_steps==max(mean_interval_steps$mean_steps),]
```
The interval with the most steps is `r max_interval[1,1]` with an average of `r max_interval[1,2]`. steps.

Checking for NA's.  I already know from my summary check above that the steps variable is the only one with any missing values.
```{r}
length(df$steps[is.na(df)])
```
I'm going to fill in the missing values with the mean steps from the corresponding time interval. 

```{r}
df_impute <- merge(df,mean_interval_steps,all.x=T) %>%
      mutate(steps = ifelse(is.na(steps),mean_steps,steps)) %>%
      select(steps,date,interval)
length(df_impute$steps[is.na(df_impute)])

```

Creating histogram of the total steps taken per day with the imputed dataset
```{r}
total_daily_steps_impute <- df_impute %>%
      group_by(date) %>%
      summarize(total_steps = sum(steps))
hist(total_daily_steps_impute$total_steps,
     main="Histogram of total daily steps with imputed data")

mean_total_daily_steps_impute <- 
      round(mean(total_daily_steps_impute$total_steps),2)
median_total_daily_steps_impute <- 
      round(median(total_daily_steps_impute$total_steps),2)
```
The mean total daily steps in the imputed data is `r mean_total_daily_steps_impute` and the median is `r median_total_daily_steps_impute` compared to `r mean_total_daily_steps` and `r median_total_daily_steps` from the original data.  Imputing the missing values results in larger mean and median estimates of total daily steps. 

Next we'll see if there is a difference in behavior between weekdays and weekends.

```{r}
# Creating factor variable to indicate weekend/weekday type
df_impute$day <- weekdays(df_impute$date)
weekend_days <- c('Saturday','Sunday')
df_impute$day_type <- as.factor(ifelse(df_impute$day %in% weekend_days,
                                       'weekend',
                                       'weekday'))
head(df_impute)
```

Make a panel plot containing a time series plot (i.e. \color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

Seeing how average steps taken by time interval differs between weekends and weekdays.
```{r}
mean_interval_steps_impute <- df_impute %>%
      group_by(day_type,interval) %>%
      summarize(mean_steps = mean(steps,na.rm=T))
head(mean_interval_steps)

ggplot(mean_interval_steps_impute, aes(x=interval,y=mean_steps)) +
      geom_line()+
      facet_wrap(~ day_type)
```